---
title: "Lab 6 - Text Mining"
link-citations: true
toc: false
---

```{r setup}
knitr::opts_chunk$set(eval = FALSE, include  = TRUE)
```

# Learning goals

- Use `unnest_tokens()` and `unnest_ngrams()` to extract tokens and n-grams from text.
- Use dplyr and ggplot2 to analyze text data

# Lab description

For this lab (due Tuesday Feb 24th at 8:30am EST) we will be working with a new text-based dataset consisting of transcribed medical reports.
The dataset contains transcription samples from [https://www.mtsamples.com/](https://www.mtsamples.com/).
We have created a (somewhat) cleaned version of the data, which can be found at [https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/mtsamples/mtsamples.csv](https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/mtsamples/mtsamples.csv).


### Setup packages

You should load in `dplyr`, `ggplot2` and `tidytext`.
If you don't already have `tidytext` then you can install with

```{r, eval=FALSE}
install.packages("tidytext")
```

### Read in Medical Transcriptions

Loading in the cleaned data from the `BIOSTAT620W26` repository:


---

## Question 1: What specialties do we have?

Use the `count()` function from `dplyr` to figure out how many different categories we have in the data.
Are these categories related? Overlapping? Evenly distributed?


---

## Question 2

- Tokenize the the words in the `transcription` column
- Count the number of times each token appears
- Visualize the top 20 most frequent words

Explain what we see from this result.
Does it makes sense? What insights (if any) do we get?

---

## Question 3

- Re-do the visualization but remove stop words before making it
- Bonus points if you remove numbers as well

What do we see know that we have removed stop words?
Does it give us a better idea of what the text is about?

---

# Question 4

Repeat question 2, but this time tokenize into bi-grams.
How does the result change if you look at tri-grams?

---

# Question 5

Using the results you got from Question 4, pick a word and count the words that appear before and after it.

---

# Question 6 

Which words are most used in each of the specialties?
You can use `group_by()` and `top_n()` from `dplyr` to have the calculations be done within each specialty.
Remember to remove stop words.
What are the 5 most-used words for each specialty?

# Question 7 - extra

Find your own insight in the data:

Ideas:

- Interesting n-grams
- See if certain words are used more in some specialties than others

