---
title: "Lab 9 - SQL"
link-citations: true
toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")
```

## Intro

In this lab, you will be working with a hospital database. There will be four tables that we will be working with: **doctor**, **visit**, **patient**, and **payment**. The main goal of the lab is to gain experience working with SQL databases. First, we will set up our SQL database:

```{r}
# install.packages(c("RSQLite", "DBI"))

library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Download tables
doctor <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/doctor.csv")
visit <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/visit.csv")
patient <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/patient.csv")
payment <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/payment.csv")

# Copy data.frames to database
dbWriteTable(con, "doctor", doctor)
dbWriteTable(con, "visit", visit)
dbWriteTable(con, "patient", patient)
dbWriteTable(con, "payment", payment)
```

```{r}
dbListTables(con)
```

TIP: You can use the following QUERY to see the structure of a table

```{sql, connection=con}
PRAGMA table_info(doctor)
```

SQL references:

[https://www.w3schools.com/sql/](https://www.w3schools.com/sql/)

# Exercise 1

Edit the code below to retrieve the doctor ID, first name and last name for all doctors using the `doctor` table. Sort by last name and then by first name (note that the code chunk below is set up to run SQL code rather than R code).

```{sql, connection=con}
SELECT 
FROM 
ORDER by 
```

# Exercise 2

Retrieve the doctor ID, first name, and last name for doctors whose last name equals 'WILLIAMS' or 'DAVIS'.

```{sql, connection=con}
SELECT 
FROM 
WHERE ___ IN ('WILLIAMS', 'DAVIS')
```

# Exercise 3

Write a query against the `visit` table that returns the IDs of the patients who visited a hospital starting on July 5, 2005 (use the visit_start_date column, and you can use the `date()` function to ignore the time component). Include a single row for each distinct patient ID. 

```{sql, connection=con}
SELECT DISTINCT 
FROM 
WHERE date(___) = '2005-07-05'
```

# Exercise 4

## Exercise 4.1

Construct a query that retrieves all columns of rows from the `payment` table where the amount is either 199, 799, or 999.

```{sql, connection=con}
SELECT *
FROM ___
WHERE ___ IN (199, 799, 999)
```

## Exercise 4.2

Construct a query that retrieves all columns of rows from the `payment` table where the amount is greater then 500.

```{sql, connection=con}
SELECT *
FROM 
WHERE 
```

## Exercise 4.2

Construct a query that retrieves all columns of rows from the `payment` table where the amount is greater then 500 and less then 800.

```{sql, connection=con}
SELECT *
FROM ___
WHERE ___ AND ___
```

# Cleanup

Run the following chunk to disconnect from the connection.

```{r}
# clean up
dbDisconnect(con)
```
